Data Specialization for a Staged Lambda Calculus

### 2014-11-05 paper/semantics.tex
minor changes to 4

More precisely, when evaluating the \verb|next|, we construct an explicit
substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
containing $\prev$ block:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
  `1`let x = (next{`2`yhat`1`}, 7) in
  next{`2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}`1`}
}`
\end{lstlisting}
%As a convention, we render the new variable with a %stylish and fashionable
%hat.  
We proceed with stage~\bbone\ evaluation, which now duplicates the variable
$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
  `1`next{`2`
    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} * 
    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} *
    hold{`1`#2 (next {`2`yhat`1`}, 7)`2`}
  `1`}`2`
}`
\end{lstlisting}
We then specialize to reduce the projections and cancel the $\next$s and $\prev$s, giving:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
    `1`next{`2`yhat * yhat * 7`1`}
`2`}`
\end{lstlisting}
Once again, we lift the results of specialization into a substitution:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
[zhat|->yhat * yhat * 7]
    `1`next{`2`zhat`1`}`2`
}`
\end{lstlisting}
Finally, when canceling the last $\next$ and $\prev$, the semantics {\em reifies} the contained substitutions into let statements, yielding
\begin{lstlisting} 
2`let yhat = f 20 in
let zhat = yhat * yhat * 7 in zhat`
\end{lstlisting}

>>>>

More precisely, when evaluating the \verb|next|, we construct an explicit
substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
containing $\prev$ block:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
  `1`let x = (next{`2`yhat`1`}, 7) in
  next{`2`prev{`1`#1 x`2`} * prev{`1`#1 x`2`} * hold{`1`#2 x`2`}`1`}`2`
}`
\end{lstlisting}
%As a convention, we render the new variable with a %stylish and fashionable
%hat.  
We proceed with stage~\bbone\ evaluation, which now duplicates the variable
$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
  `1`next{`2`
    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} * 
    prev{`1`#1 (next {`2`yhat`1`}, 7)`2`} *
    hold{`1`#2 (next {`2`yhat`1`}, 7)`2`}
  `1`}`2`
}`
\end{lstlisting}
We then specialize to reduce the projections and cancel the $\next$s and $\prev$s, giving:
\begin{lstlisting} 
2`prev {
[yhat|->f 20]
    `1`next{`2`yhat * yhat * 7`1`}
`2`}`
\end{lstlisting}
Once again, we lift the results of specialization into a substitution:
\begin{lstlisting} 
2`prev {
[yhat|->f 20,
 zhat|->yhat * yhat * 7]
    `1`next{`2`zhat`1`}`2`
}`
\end{lstlisting}
Finally, when canceling the last $\next$ and $\prev$, the semantics {\em reifies} the contained substitutions into let statements, yielding
\begin{lstlisting} 
2`let yhat = f 20 in
let zhat = yhat * yhat * 7 in zhat`
\end{lstlisting}

### 2014-11-05 paper/splitting.tex
minor changes to 4

Splitting a stage \bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
term $p$ and a \emph{resumer} function $l.r$, which contain all stage \bbone\
and \bbtwo\ subcomputations of $e$, respectively. Since stage \bbtwo\
computations can depend on its stage \bbone\ computations within $e$, 
$r$ must depend on $p$, which splitting accomplishes via the variable $l$. 
Given this form, the goal for splitting is to ensure that evaluating
$\letin l p r$ agrees with the value to which $e$ itself evaluates.

@@

Splitting stage \bbtwo\ terms, as implemented by the judgment $e \splittwosym
[p,l.r]$ (\cref{fig:termSplitTwo}), is guided by the structure of the term $e$.

\ur{At this point, I would have expected a brief discussion of the
  easy cases, like the paragraph that starts with ``In general'' below
  does. You can move that paragraph here and edit to play this role.
  You can then move onto the example below after stating that more
  interesting things happen with prev and hold.}

For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
\end{lstlisting}
has two subterms, the latter of which itself has two subterms. Two are $\pause$s
containing stage \bbone\ computations, while the third has no stage \bbone\
content.

@@

\ur{The above example should refer to Figure 11 (not Figure 10).  But
  this is not mentioned.}

\ur{This paragraph should come first, followed by a paragraph that
  talks about Figure 11.}
In general, constants and variables have no stage \bbone\ content, so their
precomputations are \texttt{()}, and their resumers return the original term's
value. Non-staging constructs with $n$ subterms have precomputations which are
$n$-tuples of their subterms' precomputations, and resumers which pass each
component to the resumer of the respective subterm. (This behavior at
\texttt{case} and functions causes precomputations to be lifted out from
underneath stage \bbtwo\ binders.) The $\prev$ and $\pause$ rules depends on
splitting stage \bbone\ terms, which is covered later.

>>>>

Splitting a stage \bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
term $p$ and a \emph{resumer} function $l.r$, which contain all stage \bbone\
and \bbtwo\ subcomputations of $e$, respectively. As stage \bbtwo\
computations within $e$ can depend on its stage \bbone\ computations, 
$r$ must depend on the results of $p$, which splitting accomplishes via the variable $l$. 
Given this form, the goal for splitting is to ensure that evaluating
$\letin l p r$ agrees with the value to which $e$ itself evaluates.

@@

Splitting stage \bbtwo\ terms, as implemented by the judgment $e \splittwosym
[p,l.r]$ (\cref{fig:termSplitTwo}), is guided by the structure of the term $e$.
For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
\end{lstlisting}
has two subterms, the latter of which itself has two subterms. Two are $\pause$s
containing stage \bbone\ computations, while the third has no stage \bbone\
content.

@@

\nr{Synopsis of Umut's comments: consider moving some of the easier rules descriptions earlier.}

In general, constants and variables have no stage \bbone\ content, so their
precomputations are \texttt{()}, and their resumers return the original term's
value. Non-staging constructs with $n$ subterms have precomputations which are
$n$-tuples of their subterms' precomputations, and resumers which pass each
component to the resumer of the respective subterm. (This behavior at
\texttt{case} and functions causes precomputations to be lifted out from
underneath stage \bbtwo\ binders.) The $\prev$ and $\pause$ rules depends on
splitting stage \bbone\ terms, and so they are covered later.
